A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-03-02 sections/Overview.tex
mechanise -> mechanize

At first sight, the DK algorithm would seem a good basis for mechanisation. After
all it comes with a careful description and extensive manual proofs.
Unfortunately, we ran into several obstacles that have prompted us to formulate
a different, more mechanisation-friendly algorithm.

@@

Besides the problematic lemmas in DK's metatheory, there are other reasons to
look for an alternative algorithmic formulation of the type system that is more
amenable to mechanisation. Indeed, two aspects that are particularly
challenging to mechanise are the scoping of universal and existential type
variables, and the propagation of the instantiation of existential type
variables across judgments. 
DK is already quite disciplined on these accounts, in particular compared to
traditional constraint-based type-inference algorithms like Algorithm $\mathcal{W}$~\cite{milner1978theory} which
features an implicit global scope for all type variables. Indeed, DK uses its
explicit and ordered context $\Gamma$ that tracks the relative scope of universal and
existential variables and it is careful to only instantiate existential
variables in a well-scoped manner.

Moreover, DK's algorithm carefully propagates instantiations by recording them
into the context and threading this context through all judgments. 
% two This means that every judgments takes two contexts---an input and an output
% context---rather than the conventional single context. The output context
% records any new variable instantations; to propagate these instantiations to
% the remaining judgments, their predecessor's output context---which is their
% input context---is applied to them as a substitution.
While this works well on paper, this approach is still fairly involved and thus
hard to reason about in a mechanised setting. Indeed, the instantiations have
to be recorded in the context and are applied incrementally to each remaining
judgment in turn, rather than immediately to all remaining judgments at once.
Also, as we have mentioned above, the stack discipline of the ordered contexts
does not mesh well with the use of output contexts; explicit marker entries are
needed in two places to demarcate the end of an existential variable's scope.
Actually we found a scoping issue related to the subsumption rule $\mathtt{DK\_Sub}$,
which might cause existential variables to leak across judgments.
A detailed discussion on this issue with a possible fix is in Section~\ref{sec:discussion:scoping}.

The complications of scoping and propagation are compelling reasons
to look for a simpler algorithm that is easier to
reason about in a mechanised setting.

@@

In recent work, \citet{itp2018} have shown how to mechanise a variant of DK's
subtyping algorithm and shown it correct with respect to DK's declarative
subtyping judgment. This approach overcomes the problem's in DK's formulation
by using a \emph{worklist}-based judgment of the form $$\Gamma \vdash \Omega$$
where $\Omega$ is a worklist (or sequence) of subtyping problems of the
form $A \leq B$.  The judgment is defined by case analysis on the first
element of $\Omega$ and recursively processes the worklist until it is empty.
Using both a single common ordered context $\Gamma$ and a worklist $\Omega$ greatly
simplifies propagating the instantiation of type variables in one
subtyping problem to the remaining ones.

>>>>

At first sight, the DK algorithm would seem a good basis for mechanization. After
all it comes with a careful description and extensive manual proofs.
Unfortunately, we ran into several obstacles that have prompted us to formulate
a different, more mechanization-friendly algorithm.

@@

Besides the problematic lemmas in DK's metatheory, there are other reasons to
look for an alternative algorithmic formulation of the type system that is more
amenable to mechanization. Indeed, two aspects that are particularly
challenging to mechanize are the scoping of universal and existential type
variables, and the propagation of the instantiation of existential type
variables across judgments. 
DK is already quite disciplined on these accounts, in particular compared to
traditional constraint-based type-inference algorithms like Algorithm $\mathcal{W}$~\cite{milner1978theory} which
features an implicit global scope for all type variables. Indeed, DK uses its
explicit and ordered context $\Gamma$ that tracks the relative scope of universal and
existential variables and it is careful to only instantiate existential
variables in a well-scoped manner.

Moreover, DK's algorithm carefully propagates instantiations by recording them
into the context and threading this context through all judgments. 
% two This means that every judgments takes two contexts---an input and an output
% context---rather than the conventional single context. The output context
% records any new variable instantations; to propagate these instantiations to
% the remaining judgments, their predecessor's output context---which is their
% input context---is applied to them as a substitution.
While this works well on paper, this approach is still fairly involved and thus
hard to reason about in a mechanized setting. Indeed, the instantiations have
to be recorded in the context and are applied incrementally to each remaining
judgment in turn, rather than immediately to all remaining judgments at once.
Also, as we have mentioned above, the stack discipline of the ordered contexts
does not mesh well with the use of output contexts; explicit marker entries are
needed in two places to demarcate the end of an existential variable's scope.
Actually we found a scoping issue related to the subsumption rule $\mathtt{DK\_Sub}$,
which might cause existential variables to leak across judgments.
A detailed discussion on this issue with a possible fix is in Section~\ref{sec:discussion:scoping}.

The complications of scoping and propagation are compelling reasons
to look for a simpler algorithm that is easier to
reason about in a mechanized setting.

@@

In recent work, \citet{itp2018} have shown how to mechanize a variant of DK's
subtyping algorithm and shown it correct with respect to DK's declarative
subtyping judgment. This approach overcomes the problem's in DK's formulation
by using a \emph{worklist}-based judgment of the form $$\Gamma \vdash \Omega$$
where $\Omega$ is a worklist (or sequence) of subtyping problems of the
form $A \leq B$.  The judgment is defined by case analysis on the first
element of $\Omega$ and recursively processes the worklist until it is empty.
Using both a single common ordered context $\Gamma$ and a worklist $\Omega$ greatly
simplifies propagating the instantiation of type variables in one
subtyping problem to the remaining ones.

### 2019-03-02 sections/Related.tex
unused comments; bad rule format; bib fix; proof script fix
mechanise -> mechanize

DK's algorithm is shown to be sound, complete and decidable in 70 pages of manual proofs.
Though carefully written, some of the proofs are incorrect
(see discussion in Appendix~\ref{appendix:false_lemmas} and \ref{appendix:subsumption}),
which creates difficulties when formalizing them in a proof assistant.
In their follow-up work \citet{DunfieldIndexed} enrich the bidirectional higher-rank system with
existentials and indexed types.
With a more complex declarative system, they developed a proof of over 150 pages.
It is even more difficult to argue its correctness for every single detail within such a big development.
Unfortunately, we find that their Lemma 26 (Parallel Admissibility) appears to have the same issue 
as lemma 29 in \citet{dunfield2013complete}: the conclusion is false. We also discuss
the issue in more detail in Appendix~\ref{appendix:false_lemmas}. 
\bruno{I think you will need to justify this better and in the appendix you must show
the lemma and counter-example.}

@@

\paragraph{Mechanical Formalizations of Other Type-Inference Algorithms}
Since the publication of the {\sc POPLMark} challenge~\cite{aydemir2005mechanized},
many theorem provers and packages provide new methods for dealing
with variable binding~\cite{aydemir2008engineering,urban2008nominalTech,chlipala2008parametric}.
More and more type systems are formalized with these tools.
However, mechanising certain algorithmic aspects, like unification and
constraint solving, has received very little attention and is still challenging.
Moreover, while most tools supports local (input) contexts in a neat way,
many practical type-inference algorithms require
more complex binding structures with output contexts or various forms of constraint solving procedures.

>>>>

DK's algorithm is shown to be sound, complete and decidable in 70 pages of manual proofs.
Though carefully written, some of the proofs are incorrect
(see discussion in Appendix~\ref{appendix:false_lemmas} and \ref{appendix:subsumption}),
which creates difficulties when formalizing them in a proof assistant.
In their follow-up work \citet{DunfieldIndexed} enrich the bidirectional higher-rank system with
existentials and indexed types.
With a more complex declarative system, they developed a proof of over 150 pages.
It is even more difficult to argue its correctness for every single detail within such a big development.
Unfortunately, we find that their Lemma 26 (Parallel Admissibility) appears to have the same issue 
as lemma 29 in \citet{dunfield2013complete}: the conclusion is false. We also discuss
the issue in more detail in Appendix~\ref{appendix:false_lemmas}.

@@

\paragraph{Mechanical Formalizations of Other Type-Inference Algorithms}
Since the publication of the {\sc POPLMark} challenge~\cite{aydemir2005mechanized},
many theorem provers and packages provide new methods for dealing
with variable binding~\cite{aydemir2008engineering,urban2008nominalTech,chlipala2008parametric}.
More and more type systems are formalized with these tools.
However, mechanizing certain algorithmic aspects, like unification and
constraint solving, has received very little attention and is still challenging.
Moreover, while most tools supports local (input) contexts in a neat way,
many practical type-inference algorithms require
more complex binding structures with output contexts or various forms of constraint solving procedures.

### 2019-03-02 sections/conclusion.tex
mechanise -> mechanize

In this paper we have provided the first mechanized formalization of type
inference for higher-ranked polymorphism. This contribution is made possible by
a new type inference algorithm for DK's declarative type system that
incorporates two novel mechanisation-friendly ideas. Firstly, we merge the
traditional type context with the recently proposed concept of judgment chains,
to accurately track variable scopes and to easily propagate substitutions.
Secondly, we use continuation-passing style to return types from the
bidirectional synthesis mode to further tasks.

>>>>

In this paper we have provided the first mechanized formalization of type
inference for higher-ranked polymorphism. This contribution is made possible by
a new type inference algorithm for DK's declarative type system that
incorporates two novel mechanization-friendly ideas. Firstly, we merge the
traditional type context with the recently proposed concept of judgment chains,
to accurately track variable scopes and to easily propagate substitutions.
Secondly, we use continuation-passing style to return types from the
bidirectional synthesis mode to further tasks.
